---
title: "Structure of rsortie"
output: rmarkdown::html_vignette
date: "January 2022"
vignette: >
  %\VignetteIndexEntry{Structure of rsortie}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rsortie)
```

## Introduction

This vignette provides an _overview_ of the **rsortie** package, with other vignettes going into specific actions and functions.  

The goal of the **rsortie** package is to support a continuous workflow from data through to model outputs, making model development, parameterization, experimentation and validation reproducible. Taking advantage of packages already developed in R, including the ability to manipulate datasets and spatial files effectively, supports larger and more complex experiments applied within SORTIE-ND. Examples of uses of **rsortie** are explored in other vignettes.

## SORTIE-ND

SORTIE-ND is an individual-based, spatially-explicit model of forest dynamics. Specifically:  

* Trees exist as individuals with x-y coordinates that interact in neighbourhoods affecting processes   
* the model operates in real space, the location of an individual affects the processes (birth, growth, death), and it is resource-dependent  


## Step 1: Set global parameters and import data  
Here we can walk through chunks of code  

## Step 2: Clean data (if needed)  

## Step 3: Update parameter files  

## Step 4: Add treatments  
See other vignettes  

## Step 5: Run the model  

## Step 6: Extract and Parse output files  
See other vignettes  
