---
title: "How to use the rsortie package"
output: rmarkdown::html_vignette
date: "December 7, 2021"
vignette: >
  %\VignetteIndexEntry{How to use the rsortie package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rsortie)
```

## Introduction

This vignette provides an _overview_ of how to use the **rSORTIE** package.  
Other vignettes go into specific actions and functions within the **rSORTIE** package.  

The goal of the **rSORTIE** package is to make the SORTIE-ND model more widely accessible to a wider audience. Operating the SORTIE-ND model through R allows for a continuous workflow, transparent and reproducible research, and clear documentation. 

SORTIE-ND is an individual tree, spatially explicit model of forest dynamics.  

* the model tracks individuals over time, trees are modelled individually, and each individual has a location and space  
* the model operates in real space, the location of an individual affects the processes (birth, growth, death), and it is resource-dependent  


## Step 1: Set global parameters and import data  
Here we can walk through chunks of code  

## Step 2: Clean data (if needed)  

## Step 3: Update parameter files  

## Step 4: Add treatments  
See other vignettes  

## Step 5: Run the model  

## Step 6: Extract and Parse output files  
See other vignettes  
